---
title: "GIS with R: examples from Slovakia"
author: Jan Povala 
date: 2019-07-26 
output:
#  beamer_presentation:
#    toc: true
  revealjs::revealjs_presentation:
    highlight: pygments
    transition: fade
---


<style type="text/css">
p { text-align: left; }
</style>


## Geospatial analysis. Why are we doing this?

1. Most of our data we collect are a recording of some phenomena that happened
   at a specific place and time.
2. Good visualisations and methods from spatio-temporal analysis field  can 
   help us understand our data.


## This tutorial

1. Spatial data and coordinate systems.
2. Introduction to `sf` package.
3. Visualisations.
4. Example of analysis: spatial autocorrelation.


## Extra resources

Recommended textbooks:

* **Geocomputation with R**. Available [online](https://geocompr.robinlovelace.net/).
* Wikle, Christopher K., Andrew Zammit-Mangion, and Noel Cressie. 2019. 
**Spatio-Temporal Statistics with R**. 1st ed. Boca Raton, Florida : CRC Press, 
[2019]: Chapman and Hall/CRC. https://doi.org/10.1201/9781351769723.

Other sources:

- [r-spatial.org](www.r-spatial.org) is a portal for people interested in
  spatial modelling.
- Packages like `sf`, `sp` tend to be well-documented, and with worked
  examples.



## Spatial data

- Data frames, but now each row has a geometry associated with it (usually a point
  or an area).
- We need to be careful with the coordinate systems:
  - If working at macro scale (at world level), we need to take into account
    curvature of the Earth. 
  - We stick to coordinate systems where we can work with Euclidean geometry.
  - The cadaster office in Slovakia uses **krovak** system, which is the
    legally binding coordinate system (coded as `EPSG:5513`).
  - This [website](https://epsg.io/5513) is useful for reading off the 
    coordinates for a particular point.


## Obtaining geometry (map) files

- The map files come in different formats. One of the most common ones are
shapefiles (`.shp`), geodatabase (`.gdb`), geopackage (`.gpk`),
geoJSON (`.geojson`). 
- This [page](https://www.geoportal.sk/sk/zbgis_smd/na-stiahnutie/) in the
[geoportal.sk](geoportal.sk) has the necessary files in different formats. We
stick to the shapefiles (`.shp`) format. The portal allows downloading the maps at
different resolutions. In this guide, we will be working with the most precise
one, which is referred to as 'Základná úroveň/ ZBGIS - Administratívne
hranice'.


## `sf` package


Install the package using:

```{r, echo=TRUE, eval=FALSE} 
install.packages('sf')
```

```{r, echo=T}
library('sf')
```

## `st_read()` to load maps

```
st_read(dsn, layer)
```

- `dsn`: file name of the map file.
- `layer`:  certain file formats allow saving multiple layers, we will work
   with single layer files (it does not mean we can't have multiple
   indicators/statistic for each location).


Load the map of Slovakia at the municipality level:
```{r, echo=T, results='hide', cache=TRUE}
mapa.obec.sf <- st_read(dsn = file.path(getwd(),
                                        'maps', 
                                        'ah_shp_0', 
                                        'obec_0.shp'))
```

## Inspecting the map object - coordinates

To see what coordinate system it use `st_crs()` function:
```{r, echo=T, cache=T}
st_crs(mapa.obec.sf)
```
Due to the missing code, we do a transformation to the official code using 
`st_transform()` function.  

```{r, echo=T, results='hide', cache=T}
crsCodeSkCadastreOffice <- 5513  # save the cadastre CRS to a constant
mapa.obec.sf <- st_transform(x = mapa.obec.sf, 
                             crs = crsCodeSkCadastreOffice)
st_crs(mapa.obec.sf)                   # 
```


## Inspecting the map object - data

```{r, cache=T}
summary(mapa.obec.sf)
```


## Inspecting the map object - data

From this, it is obvious that:

1. `IDN4` is a unique identifier for each municipality.
2. `NM4` is the name of the municipality.
3. `IDN3` is the identifier for okres.
4. `NM3` is the name for okres.
5. `IDN2` is the identifier for kraj.
5. `NM2` is the name for kraj.
7. `Shape_Leng` and `Shape_Area` are the perimeter and the area of the
   municipality.
8. It's not clear to me what `VYMERA` is given that the values slightly deviate
   from the `Shape_Area`.

The other columns are not relevant for this tutorial.

In the visualisation section, we will stick to these values even though they
are not particularly interesting. In later sections we will add more
interesting data.


## Data preparation

Before, we proceed, we rename the columns to more meaningful names:
```{r, cache=TRUE, results='hide'}
library(dplyr)
mapa.obec.sf <- mapa.obec.sf %>%
  rename(obec = NM4) %>%
  rename(obec_id = IDN4) %>%
  rename(okres = NM3) %>%
  rename(okres_id = IDN3) %>%
  rename(kraj = NM2) %>%
  rename(kraj_id = IDN2) %>%
  rename(obvod_uzemia = Shape_Leng) %>%
  rename(plocha_uzemia = Shape_Area)
```


```{r, echo=FALSE, cache=TRUE, results='hide'}
mapa.okres.sf <- st_read(dsn = file.path(getwd(), 'maps', 'ah_shp_0', 'okres_0.shp'))
mapa.okres.sf <- st_transform(x = mapa.okres.sf, crs = crsCodeSkCadastreOffice)

mapa.kraj.sf <- st_read(dsn = file.path(getwd(), 'maps', 'ah_shp_0', 'kraj_0.shp'))
mapa.kraj.sf <- st_transform(x = mapa.kraj.sf, crs = crsCodeSkCadastreOffice)

mapa.kraj.sf <- mapa.kraj.sf %>%
  rename(kraj = NM2) %>%
  rename(kraj_id = IDN2) %>%
  rename(obvod_uzemia = Shape_Leng) %>%
  rename(plocha_uzemia = Shape_Area)
mapa.okres.sf <- mapa.okres.sf %>%
  rename(okres = NM3) %>%
  rename(okres_id = IDN3) %>%
  rename(kraj = NM2) %>%
  rename(kraj_id = IDN2) %>%
  rename(obvod_uzemia = Shape_Leng) %>%
  rename(plocha_uzemia = Shape_Area)
```


## Visualisation - default `plot()` method

```{r, eval=FALSE}
plot(x = mapa.okres.sf[c('plocha_uzemia')],
     key.pos = 1) # this adds the legend to the bottom (1)
```


## Visualisation - default `plot()` method

```{r, cache=TRUE, echo=FALSE}
plot(x = mapa.okres.sf[c('plocha_uzemia')],
     key.pos = 1) # this adds the legend to the bottom (1)
```

This is a good place to start when doing exploratory analysis, but it's better
to use more sophisticated functions to produce better-looking plots.


## Visualisation - `ggplot`

`ggplot` can also be used in conjunction with spatial objects.

```{r, eval=FALSE}
library(ggplot2)  # Load the ggplot library
library(viridis)  # Load the library for a nice colour palette
ggplot(mapa.okres.sf) +
  geom_sf(aes(fill = plocha_uzemia)) +
  scale_fill_viridis("Plocha m\u00B2") +
  ggtitle("Plocha okresov") +
  theme_bw()
```


## Visualisation - `ggplot`


```{r ggplot, cache=TRUE, echo=FALSE}
library(ggplot2)  # Load the ggplot library
library(viridis)  # Load the library for a nice colour palette
ggplot(mapa.okres.sf) +
  geom_sf(aes(fill = plocha_uzemia)) +
  scale_fill_viridis("Plocha m\u00B2") +
  ggtitle("Plocha okresov") +
  theme_bw()
```


## Visualisation - `tmap`

This is a very versatile library that allows integration with Open Street Map
data and interactive maps on top of regular plotting. What we show here is only
scratching the surface, for more examples, please visit the [tmap's github
page](https://github.com/mtennekes/tmap).

Before we proceed, we install the libraries and load them.

```{r, eval=FALSE, echo=TRUE, results='hide', message=F, warning=F}
install.packages('tmap')
```

```{r, echo=TRUE, cache=TRUE, results='hide'}
library(tmap)
```


## Visualisation - `tmap`

```{r, eval=TRUE, CACHE=FALSE, results='hide', echo=FALSE} 
tmap_mode("plot")
```

```{r, eval=FALSE}
tmap_mode("plot")
tm_shape(mapa.okres.sf) +
  tm_fill(col = "plocha_uzemia", style = "cont", title="Plocha m\u00B2") + 
  tm_style('gray', title = 'Plocha okresov m\u00B2')
```


## Visualisation - `tmap`

```{r, cache=TRUE, echo=FALSE}
tm_shape(mapa.okres.sf) +
  tm_fill(col = "plocha_uzemia", style = "cont", title="Plocha m\u00B2") + 
  tm_style('gray', title = 'Plocha okresov m\u00B2')
```


## Visualisation - `tmap`

For example, we can do a breakdown by kraj.

```{r, eval=FALSE}
tm_shape(mapa.okres.sf) +
  tm_fill(col = "plocha_uzemia", style = "cont", title="Plocha m\u00B2") + 
  tm_style('gray', title = 'Okresy SR') + 
  tm_facets('kraj')
```


## Visualisation - `tmap`

```{r, cache=TRUE, echo=FALSE}
tm_shape(mapa.okres.sf) +
  tm_fill(col = "plocha_uzemia", style = "cont", title="Plocha m\u00B2") + 
  tm_style('gray', title = 'Okresy SR') + 
  tm_facets('kraj')
```


## Visualisation - `tmap`

For plots with fewer components (e.g. kraj), we might want to add labels. Be
careful with the overlaps of the labels and overflows outside the frame. [This
post](https://www.jla-data.net/eng/adjusting-bounding-box-of-a-tmap-map/) gives
a good solution for the latter.

```{r, eval=FALSE}
tm_shape(mapa.kraj.sf) +
  tm_fill(col = "plocha_uzemia", style = "cont", title="Plocha m\u00B2") + 
  tm_style('gray', title = 'Okresy SR') + 
  tm_text('kraj', auto.placement = TRUE)
```

To save the map, use `tmap_save()` function.


## Visualisation - `tmap`

```{r, cache=TRUE, echo=FALSE}
tm_shape(mapa.kraj.sf) +
  tm_fill(col = "plocha_uzemia", style = "cont", title="Plocha m\u00B2") + 
  tm_style('gray', title = 'Okresy SR') + 
  tm_text('kraj', auto.placement = TRUE)
```


```{r, eval=F}
mapa.okres.sf.gps <- st_transform(x = mapa.okres.sf, crs=4326)
osm_tiles = tmaptools::read_osm(st_bbox(mapa.okres.sf.gps), type = "osm")
tmap.object <- qtm(osm_tiles, raster.alpha = 0.6) + 
  tm_shape(mapa.okres.sf.gps) +
  tm_fill(col = "Shape_Area", style = "cont", title="Plocha m\u00B2") + 
  tm_style('gray', title = 'Okresy SR')
```


## Case study: presidential elections 2019

Load the data:
```{r load_xlsx_prezidentske, results='hide', echo=TRUE, message=F, warning=F}
library(readxl)
results.obec.file.path <- file.path(getwd(), 'data', "raw", 
                                    "PRE_2019_KOLO1_xlsx",
                                    "PRE_2019_KOLO1_tab03e.xlsx")
results.obec <- read_excel(results.obec.file.path)
results.obec <- results.obec[-(1:4), ]
names(results.obec) <- c('kraj_id', 'kraj', 'uzemny_obvod_id', 'uzemny_obvod',
                         'okres_id', 'okres', 'obec_id', 'obec',
                         'candidate_id', 'candidate_fname', 'candidate_sname',
                         'platne_hlasy_count', 'platne_hlasy_percent',
                         'pullouts')
results.obec <- results.obec %>%
  dplyr::mutate_at(c("platne_hlasy_count", "platne_hlasy_percent"), as.numeric)
```


## Case study: presidential elections 2019

Join the data with the map:
```{r prezidentke_map_join, echo=T, results='hide', message=F, warning=F}
results.obec.sf <- mapa.obec.sf %>%
  dplyr::select(obec_id) %>%
  dplyr::mutate(obec_id = as.character(obec_id)) %>%
  dplyr::full_join(results.obec, by = c("obec_id" = "obec_id")) %>%
  dplyr::mutate(name = paste(candidate_fname, candidate_sname, sep = " "))
```


## Case study: presidential elections 2019

Since we are working with 'obec' level, we focus only on one 'kraj' region, and
we only compare top 4 candidates in that region.
```{r, cache=T, echo=T, results='hide', message=F, warning=F}
zilinsky.kraj.results.sf <- results.obec.sf %>%
  dplyr::filter(kraj_id == 5)  # zilinsky kraj

# select top 4 candidates in Zilinsky kraj
zilinsky.kraj.top.candidates  <- zilinsky.kraj.results.sf %>%
  dplyr::group_by(name) %>%
  dplyr::summarise(vote_total = sum(platne_hlasy_count)) %>%
  dplyr::arrange(vote_total) %>%
  dplyr::top_n(4, vote_total) %>%
  dplyr::pull(name)

# filter out only the 4 candidates
zilinsky.kraj.results.top.candidates.sf <- zilinsky.kraj.results.sf %>%
  dplyr::filter(name %in% zilinsky.kraj.top.candidates)
```


## Case study: presidential elections 2019

```{r, eval=F, echo=T}
tmap_mode("plot")
tm_shape(zilinsky.kraj.results.top.candidates.sf) +
  tm_fill(col = "platne_hlasy_percent", style = "cont", title="Podiel hlasov %") + 
  tm_style('gray', title = 'Žilinsky kraj') + 
  tm_facets('name')
```


## Case study: presidential elections 2019

```{r, cache=T, message=F, warning=F, echo=F}
tmap_mode("plot")
tm_shape(zilinsky.kraj.results.top.candidates.sf) +
  tm_fill(col = "platne_hlasy_percent", style = "cont", title="Podiel hlasov %") + 
  tm_style('gray', title = 'Žilinsky kraj') + 
  tm_facets('name')
```


## Case study: presidential elections 2019

As we can see this visualisation is not ideal, so we go for the interactive
map where we can zoom in/out as we please. We only focus on Zuzana Caputova
and Maros Sefcovic.

```{r, eval=F, echo=T}
library(tidyr)

zilinsky.kraj.results.top.2.sf <- zilinsky.kraj.results.top.candidates.sf %>%
  dplyr::filter(name %in% c('Maroš Šefčovič', 'Zuzana Čaputová'))

zilinsky.results.wide.sf <- zilinsky.kraj.results.top.2.sf %>%
  dplyr::select(name, platne_hlasy_percent, obec) %>%
  tidyr::spread(name, platne_hlasy_percent)

tmap_mode("view")
tm <- tm_shape(zilinsky.results.wide.sf) + 
  tm_polygons(c('Maroš Šefčovič', 'Zuzana Čaputová')) + 
  tm_facets(sync = TRUE, ncol = 2)
tm
```


## Case study: presidential elections 2019
```{r prezidentske_interactive, cache=F, echo=F, message=F, warning=F}
library(tidyr)

zilinsky.kraj.results.top.2.sf <- zilinsky.kraj.results.top.candidates.sf %>%
  dplyr::filter(name %in% c('Maroš Šefčovič', 'Zuzana Čaputová'))

zilinsky.results.wide.sf <- zilinsky.kraj.results.top.2.sf %>%
  dplyr::select(name, platne_hlasy_percent, obec) %>%
  tidyr::spread(name, platne_hlasy_percent)

tmap_mode("view")
tm <- tm_shape(zilinsky.results.wide.sf) + 
  tm_polygons(c('Maroš Šefčovič', 'Zuzana Čaputová')) + 
  tm_facets(sync = TRUE, ncol = 2) +
  tm_layout(legend.text.size = 0.4)
tm

```
