---
title: "GIS with R: examples from Slovakia"
date: July 2019
output:
  revealjs::revealjs_presentation:
    highlight: pygments
    transition: fade
---

## Geospatial analysis. Why are we doing this?

<style type="text/css">

/* Headers */
h1,h2,h3,h4,h5,h6{
  font-size: 20pt;
}

code.r{
  font-size: 12px;
}
pre {
  font-size: 12px;
}

p {
  text-align: left;
}
</style>

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 120)
```

1. Most of our data we collect are a recording of some phenomena that happened
   at a specific place and time.
2. Good visualisations and methods from spatio-temporal analysis field  can 
   help us understand our data.


## This tutorial

1. Spatial data and coordinate systems.
2. Introduction to `sf` package.
3. Visualisations.
4. Example of analysis: spatial autocorrelation.


## Extra resources

Recommended textbooks:

<small>

* **Geocomputation with R**. Available [online](https://geocompr.robinlovelace.net/).
* Gelfand, Alan E., ed. 2010. **Handbook of Spatial Statistics**. Chapman & Hall/CRC Handbooks 
  of Modern Statistical Methods. Boca Raton: CRC Press. 
* Wikle, Christopher K., Andrew Zammit-Mangion, and Noel Cressie. 2019. 
**Spatio-Temporal Statistics with R**. 1st ed. Boca Raton, Florida : CRC Press, 
[2019]: Chapman and Hall/CRC. https://doi.org/10.1201/9781351769723.

</small>

Other sources:
<small>

- [r-spatial.org](www.r-spatial.org) is a portal for people interested in
  spatial modelling.
- Packages like `sf`, `sp` tend to be well-documented, and with worked
  examples.

</small>

## Spatial data

- Data frames, but now each row has a geometry associated with it (usually a point
  or an area).
- We need to be careful with the coordinate systems:
  - If working at macro scale (at world level), we need to take into account
    curvature of the Earth. 
  - We stick to coordinate systems where we can work with Euclidean geometry.
  - The cadaster office in Slovakia uses **krovak** system, which is the
    legally binding coordinate system (coded as `EPSG:5513`).
  - This [website](https://epsg.io/5513) is useful for reading off the 
    coordinates for a particular point.


## Obtaining geometry (map) files

- The map files come in different formats. One of the most common ones are
shapefiles (`.shp`), geodatabase (`.gdb`), geopackage (`.gpk`),
geoJSON (`.geojson`). 
- This [page](https://www.geoportal.sk/sk/zbgis_smd/na-stiahnutie/) in the
[geoportal.sk](geoportal.sk) has the necessary files in different formats. We
stick to the shapefiles (`.shp`) format. The portal allows downloading the maps at
different resolutions. In this guide, we will be working with the most precise
one, which is referred to as 'Základná úroveň/ ZBGIS - Administratívne
hranice'.


## `sf` package


Install the package using:

```{r, echo=TRUE, eval=FALSE} 
install.packages('sf')
```

```{r, echo=T}
library('sf')
```

## `st_read()` to load maps

```
st_read(dsn, layer)
```

- `dsn`: file name of the map file.
- `layer`:  certain file formats allow saving multiple layers, we will work
   with single layer files (it does not mean we can't have multiple
   indicators/statistic for each location).


Load the map of Slovakia at the municipality level:
```{r, echo=T, results='hide', cache=TRUE}
mapa.obec.sf <- st_read(dsn = file.path(getwd(),
                                        'maps', 
                                        'ah_shp_0', 
                                        'obec_0.shp'))
```

## Inspecting the map object - coordinates

To see what coordinate system it uses,  `st_crs()` function can be used:
```{r, echo=T, cache=T}
st_crs(mapa.obec.sf)
```
Due to the missing code, we do a transformation to the official code using 
`st_transform()` function.  

```{r, echo=T, results='hide', cache=T}
crsCodeSkCadastreOffice <- 5513  # save the cadastre CRS to a constant
mapa.obec.sf <- st_transform(x = mapa.obec.sf, 
                             crs = crsCodeSkCadastreOffice)
st_crs(mapa.obec.sf)
```


## Inspecting the map object - data

```{r, cache=T}
summary(mapa.obec.sf)
```


## Inspecting the map object - data

<small>

From the columns included in the shapefiles, we see that:

1. `IDN4` is a unique identifier for each municipality.
2. `NM4` is the name of the municipality.
3. `IDN3` is the identifier for okres.
4. `NM3` is the name for okres.
5. `IDN2` is the identifier for kraj.
5. `NM2` is the name for kraj.
7. `Shape_Leng` and `Shape_Area` are the perimeter and the area of a municipality.
8. It's not clear to me what `VYMERA` is given that the values slightly deviate
   from the `Shape_Area`.

The other columns are not relevant for this tutorial.

</small>


## Data preparation

Before, we proceed, we rename the columns to more meaningful names:
```{r, cache=TRUE, results='hide'}
library(dplyr)
mapa.obec.sf <- mapa.obec.sf %>%
  rename(obec = NM4) %>%
  rename(obec_id = IDN4) %>%
  rename(okres = NM3) %>%
  rename(okres_id = IDN3) %>%
  rename(kraj = NM2) %>%
  rename(kraj_id = IDN2) %>%
  rename(obvod_uzemia = Shape_Leng) %>%
  rename(plocha_uzemia = Shape_Area)
```


```{r, echo=FALSE, cache=TRUE, results='hide'}
mapa.okres.sf <- st_read(dsn = file.path(getwd(), 'maps', 'ah_shp_0', 'okres_0.shp'))
mapa.okres.sf <- st_transform(x = mapa.okres.sf, crs = crsCodeSkCadastreOffice)

mapa.kraj.sf <- st_read(dsn = file.path(getwd(), 'maps', 'ah_shp_0', 'kraj_0.shp'))
mapa.kraj.sf <- st_transform(x = mapa.kraj.sf, crs = crsCodeSkCadastreOffice)

mapa.kraj.sf <- mapa.kraj.sf %>%
  rename(kraj = NM2) %>%
  rename(kraj_id = IDN2) %>%
  rename(obvod_uzemia = Shape_Leng) %>%
  rename(plocha_uzemia = Shape_Area)
mapa.okres.sf <- mapa.okres.sf %>%
  rename(okres = NM3) %>%
  rename(okres_id = IDN3) %>%
  rename(kraj = NM2) %>%
  rename(kraj_id = IDN2) %>%
  rename(obvod_uzemia = Shape_Leng) %>%
  rename(plocha_uzemia = Shape_Area)
```

## Visualisation - options

Some of the most popular options are:

1. default `plot()` method in the `sf` package,
2. `ggplot`,
3. `tmap`.

We will give examples of all throughout the rest of this tutorial.


## Example: migration in Slovakia

1. Load the migration and population data.
2. Merge the data with the map files.
3. Visualise the data: inflow / outflow. 


## Example: migration in Slovakia
```{r, echo=F}
library("readxl")
```
Load migration and population data
```{r migration_data, cache=T, warning=F, message=F, results='hide'}
migration.district.df <- read_excel(file.path(getwd(), 'data', "raw", "migration_by_district.xlsx"),
                                    skip=8, col_names=FALSE)
colnames(migration.district.df) <- c('district', 'inflow_sk', 'inflow_abroad', 'outflow_sk', 'outflow_abroad')
migration.district.df <- migration.district.df %>%
  dplyr::mutate(district = gsub("^District of (.+)$", "\\1", district))

# load population data and clean them
population.municipality.df <- read_excel(file.path(getwd(), 'data', "raw", "population_municipality.xlsx"),
                                         skip=7, col_names=FALSE)
colnames(population.municipality.df) <- c('district', 'municipality', 'population')
population.district.df <- population.municipality.df  %>%
  dplyr::mutate(district = gsub("^District of (.+)$", "\\1", district)) %>%
  dplyr::group_by(district) %>%
  dplyr::summarise(population = sum(population))
```


## Example: migration in Slovakia

Merge the data with maps
```{r, cache=T, warning=F, message=F, results='hide'}
library(stringi) # for removing diacritics for joins
pop.migration.district.df <- migration.district.df %>%
  dplyr::full_join(population.district.df, by = 'district') %>%
  dplyr::mutate(district_ascii = stri_trans_general(str = district, id = "Latin-ASCII"))

pop.migration.district.sf <- mapa.okres.sf %>%
  dplyr::mutate(district_ascii = stri_trans_general(str = okres, id = "Latin-ASCII")) %>%
  dplyr::left_join(pop.migration.district.df, by = "district_ascii")

pop.migration.district.sf <- pop.migration.district.sf %>%
  dplyr::mutate(outflow_abroad_percent = 100*(outflow_abroad / population)) %>%
  dplyr::mutate(outflow_sk_percent = 100*(outflow_sk / population)) %>%
  dplyr::mutate(inflow_abroad_percent = 100*(inflow_abroad / population)) %>%
  dplyr::mutate(inflow_sk_percent = 100*(inflow_sk / population))
```


## Example: migration in Slovakia

Plot using the default `plot()` method
```{r, echo=T, eval=F}
plot(x = pop.migration.district.sf[c('outflow_abroad_percent')], 
     key.pos = 1, main='% of population moved abroad')
```


## Example: migration in Slovakia

```{r, echo=F, cache=T, warning=F, message=F, fig.align = 'center'}
plot(x = pop.migration.district.sf[c('outflow_abroad_percent')], 
     key.pos = 1, main='% of population moved abroad')
```


## Example: migration in Slovakia

Plotting side-by-side using the default method is possible, but has
limitations such as not supporting legends. `tmap` package can handle those
very well.

`tmap` allows integration with Open Street Map
data and interactive maps on top of regular plotting. What we show here is only
scratching the surface, for more examples, please visit the [tmap's github
page](https://github.com/mtennekes/tmap).

```{r, eval=FALSE, echo=TRUE, results='hide', message=F, warning=F}
install.packages('tmap')
```


```{r, echo=TRUE, cache=TRUE, results='hide'}
library(tmap)
```


## Example: migration in Slovakia

```{r, eval=F}
pop.migration.sf.tall <- pop.migration.district.sf %>%
  tidyr::gather("type", "percent", c("outflow_abroad_percent", "outflow_sk_percent",
                                     "inflow_abroad_percent", "inflow_sk_percent")) %>%
  dplyr::mutate(type = gsub("_", " ", gsub("(.+)\\_percent", "\\1", type)))

sk.migration <- pop.migration.sf.tall %>%
  dplyr::filter(grepl('.*sk', type))

tm_shape(sk.migration) +
  tm_fill(col = "percent", style = "cont", title="% of population") + 
  tm_style('gray', title = 'Migration within Slovakia') + 
  tm_facets('type')
```

Exercise: do the same for migration to/out of Slovakia.


## Example: migration in Slovakia

```{r, echo=F, cache=T, warning=F, message=F, fig.align = 'center'}
pop.migration.sf.tall <- pop.migration.district.sf %>%
  tidyr::gather("type", "percent", c("outflow_abroad_percent", "outflow_sk_percent",
                                     "inflow_abroad_percent", "inflow_sk_percent")) %>%
  dplyr::mutate(type = gsub("_", " ", gsub("(.+)\\_percent", "\\1", type)))

sk.migration <- pop.migration.sf.tall %>%
  dplyr::filter(grepl('.*sk', type))

tm_shape(sk.migration) +
  tm_fill(col = "percent", style = "cont", title="% of population") + 
  tm_style('gray', title = 'Migration within Slovakia') +
  tm_facets('type')
```


## Case study: respiratory diseases

1. Download the dataset from the [official statistics portal](http://datacube.statistics.sk/#!/view/sk/VBD_DEM/om7035rr/Zomrel%C3%AD%20pod%C4%BEa%20pr%C3%AD%C4%8Diny%20smrti,%20pohlavia%20a%20trval%C3%A9ho%20pobytu%20-%20SR-oblas%C5%A5-kraj-okres,%20m-v%20%5Bom7035rr%5D)
2. Plot the rate of deaths due to respiratory diseases.
3. Is there spatial correlation?


## Case study: respiratory diseases

```{r, cache=T, warning=F, message=F, results='hide'}
death.causes.df <- read_excel(file.path(getwd(), 'data', "raw", "deaths_causes.xlsx"), skip=7, col_names=FALSE)
colnames(death.causes.df) <- c('district', 'group', 'type', 'y2014', 'y2015', 'y2016', 'y2017', 'y2018')

# sum over the years, merge with population counts and group by disease group
respiratory.causes.df <- death.causes.df %>%
  dplyr::mutate(district = gsub("^District of (.+)$", "\\1", district)) %>%
  dplyr::mutate(counts_5year = (y2014 + y2015 + y2016 + y2017 + y2018) / 5) %>%
  dplyr::group_by(district, group) %>%
  dplyr::summarise(death_count = sum(counts_5year)) %>%
  dplyr::filter(group == 'X Diseases of the respiratory system') %>%
  dplyr::full_join(population.district.df, by = 'district') %>%
  dplyr::mutate(death_rate = 100 * (death_count / population)) %>%
  dplyr::mutate(district_ascii = stri_trans_general(str = district, id = "Latin-ASCII"))

# merge with the map
respiratory.causes.sf <- mapa.okres.sf %>%
  dplyr::mutate(district_ascii = stri_trans_general(str = okres, id = "Latin-ASCII")) %>%
  dplyr::left_join(respiratory.causes.df, by = "district_ascii")
```


## Case study: respiratory diseases

For a change, we will use `ggplot`:

```{r, eval=F}
library(ggplot2)  # Load the ggplot library
library(viridis)  # Load the library for a nice colour palette
ggplot(respiratory.causes.sf) +
  geom_sf(aes(fill = death_rate)) +
  scale_fill_viridis("Death rate") +
  ggtitle("Deaths due to respiratory diseases") +
  theme_bw()
```


## Case study: respiratory diseases

```{r, echo=F, cache=T, fig.align='center'}
library(ggplot2)  # Load the ggplot library
library(viridis)  # Load the library for a nice colour palette
ggplot(respiratory.causes.sf) +
  geom_sf(aes(fill = death_rate)) +
  scale_fill_viridis("Death rate") +
  ggtitle("Deaths due to respiratory diseases") +
  theme_bw()
```


## Case study: respiratory diseases

* The plot suggests that districts next to each other often have very similar
  rate (btw. the hotspot is Turčianske Teplice).
* A common test for the presence of spatial autocorrelation in spatial 
  statistics is so called _Moran's I test_.
* A value very close to zero means that there is no spatial correlation
  present.

$$
\begin{equation}
I=\frac{m \sum_{i=1}^{m} \sum_{j=1}^{m} w_{i j}\left(Z_{i}-\overline{Z}\right)\left(Z_{j}-\overline{Z}\right)}{\left(\sum_{i=1}^{m} \sum_{j=1}^{m} w_{i j}\right)\left(\sum_{i=1}^{m}\left(Z_{i}-\overline{Z}\right)^{2}\right)}
\end{equation}
$$


## Case study: respiratory diseases

```{r moran, cache=T}
library(spatstat)
library(sp)
library(spdep)

death.rates <- unlist(st_drop_geometry(respiratory.causes.sf['death_rate']))
nb.list <- poly2nb(pl = respiratory.causes.sf, snap = 1)
nb.weights <- nb2listw(nb.list) 
spdep::moran.test(death.rates, nb.weights) 
```

## Case study: respiratory diseases

<small>

* The value of Moran's I statistic is 0.177 which is significantly above 0. The
  p-value is 0.003993.
* This suggests that the differences in the death rates are not only due to
  chance. This gives us indication to explore further and find explanatory
  variables that would explain the differences.
* **Challenge exercise:** take some explanatory variable (e.g. altitude, level of
  pollution), fit a simple linear model, and check the spatial autocorrelation
  of the residuals. Hopefully, adding explanatory variables reduces it.
* Often, we are not able to observe the relevant covariates. To account for
  this unexplained variation, a latent component is often introduced. Often it
  is a Gaussian random field. When the field is assumed to be Markovian, 
  these can be fit efficiently. This is beyond the scope of the course, but
  models to watch out for are: CAR Models, BYM Models (refinement of CAR),
  log-Gaussian Cox process (for count data).

</small>

## Case study: elections 2019

Next, we look at presidential elections in Slovakia from 2019.

1. Download the raw dataset of the results from [here](http://volby.statistics.sk/prez/prez2019/sk/download.html).
2. Load the results, merge it with the geography you are interested in (obec, kraj)
3. Plot the results, including interactive maps.


## Case study: elections 2019

Load the data:
```{r load_xlsx_prezidentske, results='hide', echo=TRUE, message=F, warning=F}
library(readxl)
results.obec.file.path <- file.path(getwd(), 'data', "raw", 
                                    "PRE_2019_KOLO1_xlsx",
                                    "PRE_2019_KOLO1_tab03e.xlsx")
results.obec <- read_excel(results.obec.file.path)
results.obec <- results.obec[-(1:4), ]
names(results.obec) <- c('kraj_id', 'kraj', 'uzemny_obvod_id', 'uzemny_obvod',
                         'okres_id', 'okres', 'obec_id', 'obec',
                         'candidate_id', 'candidate_fname', 'candidate_sname',
                         'platne_hlasy_count', 'platne_hlasy_percent',
                         'pullouts')
results.obec <- results.obec %>%
  dplyr::mutate_at(c("platne_hlasy_count", "platne_hlasy_percent"), as.numeric)
```


## Case study: elections 2019

Join the data with the map:
```{r prezidentke_map_join, echo=T, results='hide', message=F, warning=F}
results.obec.sf <- mapa.obec.sf %>%
  dplyr::select(obec_id) %>%
  dplyr::mutate(obec_id = as.character(obec_id)) %>%
  dplyr::full_join(results.obec, by = c("obec_id" = "obec_id")) %>%
  dplyr::mutate(name = paste(candidate_fname, candidate_sname, sep = " "))
```


## Case study: elections 2019

Since we are working with 'obec' level, we focus only on one 'kraj' region, and
we only compare top 4 candidates in that region.
```{r, cache=T, echo=T, results='hide', message=F, warning=F}
zilinsky.kraj.results.sf <- results.obec.sf %>%
  dplyr::filter(kraj_id == 5)  # zilinsky kraj

# select top 4 candidates in Zilinsky kraj
zilinsky.kraj.top.candidates  <- zilinsky.kraj.results.sf %>%
  dplyr::group_by(name) %>%
  dplyr::summarise(vote_total = sum(platne_hlasy_count)) %>%
  dplyr::arrange(vote_total) %>%
  dplyr::top_n(4, vote_total) %>%
  dplyr::pull(name)

# filter out only the 4 candidates
zilinsky.kraj.results.top.candidates.sf <- zilinsky.kraj.results.sf %>%
  dplyr::filter(name %in% zilinsky.kraj.top.candidates)
```


## Case study: elections 2019

```{r, eval=F, echo=T}
tmap_mode("plot")
tm_shape(zilinsky.kraj.results.top.candidates.sf) +
  tm_fill(col = "platne_hlasy_percent", style = "cont", title="Podiel hlasov %") + 
  tm_style('gray', title = 'Žilinsky kraj') + 
  tm_facets('name')
```


## Case study: elections 2019

```{r, cache=T, message=F, warning=F, echo=F, fig.align = 'center'}
tmap_mode("plot")
tm_shape(zilinsky.kraj.results.top.candidates.sf) +
  tm_fill(col = "platne_hlasy_percent", style = "cont", title="Podiel hlasov %") + 
  tm_style('gray', title = 'Žilinsky kraj') + 
  tm_facets('name')
```


## Case study: elections 2019

As we can see this visualisation is not ideal, so we go for  an interactive
map where we can zoom in/out as we please.

```{r, eval=F, echo=T}
library(tidyr)

zilinsky.kraj.results.top.2.sf <- zilinsky.kraj.results.top.candidates.sf %>%
  dplyr::filter(name %in% c('Maroš Šefčovič', 'Zuzana Čaputová'))

zilinsky.results.wide.sf <- zilinsky.kraj.results.top.2.sf %>%
  dplyr::select(name, platne_hlasy_percent, obec) %>%
  tidyr::spread(name, platne_hlasy_percent)

tmap_mode("view")
tm <- tm_shape(zilinsky.results.wide.sf) + 
  tm_polygons(c('Maroš Šefčovič', 'Zuzana Čaputová')) + 
  tm_facets(sync = TRUE, ncol = 2) +
  tm_layout(legend.text.size = 0.4)
tm
```


## Case study: elections 2019

```{r fig.width=6, fig.height=6, fig.align = 'center', cache=T,echo=FALSE}
library(png)
library(grid)
img <- readPNG("./images/elections-zilinsky-interactive.png")
 grid.raster(img)
```
