---
title: "GIS with R: tutorial with focus on Slovakia"
output:
    html_document:
        keep_md: true
---

# Useful resources

1. Geocomputation with R: a comprehensive, but very practical book on GIS with
   R. It is freely available online at [this
   link](https://geocompr.robinlovelace.net/).
2. [r-spatial.org](r-spatial.org)


# Installing `sf` package
Install the package using:
```{r, echo=TRUE, eval=FALSE}
install.packages(sf)
```

The installation should work out of the box, but one might get compatibility
issues with the `gdal` library. Depending on the error, googling it should do
the job.

After the installation, we simply load the package, and also set the R working
directory appropriately.

```{r, echo=T, results='hide', cache=TRUE}
library(sf)

setwd(file.path(path.expand('~'), 'workplace', 'gis-slovakia'))
```

# Maps

In order to be able to work with maps, we need obtain the files that have the
geometrical representation of the maps, with the corresponding coordinate
system. When working with global data, it is useful to use a coordinate system
which takes into account the curvature of the Earth, but for analysis that
involves small area (like Slovakia), it is practical to worth with planar
coordinates where we can use Euclidean geometry that we are familiar with. In
our analysis we will use the **krovak** coordinate system which covers Czechia
and Slovak Republic. The legally-binding system that the cadastre uses is coded
as `EPSG:5513`.

This [website](https://epsg.io/5513) is useful for navigating the map and
reading off the coordinates for a particular point. To change the coordinate
system, change the coordinates system code. 

## File formats
The map files come in different formats. One of the most common ones are
shapefiles (`.shp`), geodatabase (`.gdb`), geopackage (`.gpk`),
geoJSON (`.geojson`). 

## Obtaining Slovakia maps

This [page](https://www.geoportal.sk/sk/zbgis_smd/na-stiahnutie/) in the
[geoportal.sk](geoportal.sk) has the necessary files in different formats. We
stick to the shapefiles (`.shp`). The portal allows downloading the maps at
different resolutions. In this guide, we will be working with the most precise
one, which is referred to as 'Základná úroveň/ ZBGIS - Administratívne
hranice'.


# Loading the maps

`sf` package provides `st_read()` function with two parameters:
1. `dsn`: this stands for data source name, but it is effectively a file name.
2. `layer `: certain file formats allow saving multiple layers, we will work
   with single layer files (it does not mean we can't have multiple
   indicators/statistic for each location).

We start off with the maps at the municipality level. We simply run:

```{r, echo=T}
mapa.obec.sf <- st_read(dsn = file.path(getwd(), 'maps', 'ah_shp_0', 'obec_0.shp'))
```

Now that, we loaded the object, we familiarise ourselves with it.


## Coordinates system
To see what coordinate system it uses, we run:

```{r, echo=T}
st_crs(mapa.obec.sf)
```

We see that the CRS used is `krovak`, but we do not get the code, e.g. 5513.
Since, the krovak CRS comes in different flavours, we do a transformation to
the one that the cadastre office in Slovakia uses by using `st_transform()`
function.  We also save that code into a constant for future use.

```{r, echo=T}
CRS.CODE.KU.SK <- 5513    # save the cadastre CRS to a constant
mapa.obec.sf <- st_transform(x = mapa.obec.sf, crs = CRS.CODE.KU.SK)
st_crs(mapa.obec.sf)                   # 
```

Now, we see that the EPSG code for our coordinates system is set correctly.


## Data associated with the maps

Now, we explore the data that the maps comes with. First we check the column
names:

```{r}
names(mapa.obec.sf)
```

Some of these names are intuitive, but have a one more peek at the data to
understand more. R's built-in command `summary()` is a useful start.

```{r}
summary(mapa.obec.sf)
```

From this, it is obvious that:
1. `IDN4` is a unique identifier for each municipality.
2. `NM4` is the name of the municipality.
3. `IDN3` is the identifier for okres.
4. `NM3` is the name for okres.
5. `IDN2` is the identifier for kraj.
5. `NM2` is the name for kraj.
7. `Shape_Leng` and `Shape_Area` are the perimeter and the area of the
   municipality.
8. It's not clear to me what `VYMERA` is given that the values slightly deviate
   from the `Shape_Area`.

The other columns are not relevant for this tutorial.

In the visualisation section, we will stick to these values even though they
are not particularly interesting. In later sections we will add more
interesting data.

Before, we proceed, we rename them to more meaningful names:

```{r, cache=TRUE}
library(dplyr)
mapa.obec.sf <- mapa.obec.sf %>%
  rename(obec = NM4) %>%
  rename(obec_id = IDN4) %>%
  rename(okres = NM3) %>%
  rename(okres_id = IDN3) %>%
  rename(kraj = NM2) %>%
  rename(kraj_id = IDN2) %>%
  rename(obvod_uzemia = Shape_Leng) %>%
  rename(plocha_uzemia = Shape_Area)
```



## Maps for other geometries

For convenience, we also load other geometries: okres, kraj. We also transform
them into the correct CRS.

```{r, cache=TRUE, results='hide', echo=TRUE}
mapa.okres.sf <- st_read(dsn = file.path(getwd(), 'maps', 'ah_shp_0', 'okres_0.shp'))
mapa.okres.sf <- st_transform(x = mapa.okres.sf, crs = CRS.CODE.KU.SK)

mapa.kraj.sf <- st_read(dsn = file.path(getwd(), 'maps', 'ah_shp_0', 'kraj_0.shp'))
mapa.kraj.sf <- st_transform(x = mapa.kraj.sf, crs = CRS.CODE.KU.SK)
```

And we do the renaming as before (not shown here).

```{r, echo=FALSE, cache=TRUE, results='hide'}
mapa.kraj.sf <- mapa.kraj.sf %>%
  rename(kraj = NM2) %>%
  rename(kraj_id = IDN2) %>%
  rename(obvod_uzemia = Shape_Leng) %>%
  rename(plocha_uzemia = Shape_Area)
mapa.okres.sf <- mapa.okres.sf %>%
  rename(okres = NM3) %>%
  rename(okres_id = IDN3) %>%
  rename(kraj = NM2) %>%
  rename(kraj_id = IDN2) %>%
  rename(obvod_uzemia = Shape_Leng) %>%
  rename(plocha_uzemia = Shape_Area)
```


# Visualisations

One of the mosti important aspects of GIS is data visualisation. There  are
different way of going about it, here is a selection of what I've used
personally.

## Default method

Instinctively, the first thing one does when exploring an unknown object in R
is call the `plot()` method. Indeed, there is a default method that does
surprisingly good job with little effort.

```{r, cache=TRUE, echo=TRUE}
plot(mapa.okres.sf)
```

The command tried to plot all the columns, but by default it takes the first 9. 
Usually, we want to have more control over what we plot, so we choose what we
are interested in. In this case, we plot the area and the perimeter.
```{r, cache=TRUE, echo=TRUE}
plot(x = mapa.okres.sf[c('obvod_uzemia', 'plocha_uzemia')],
     key.pos = 1) # this add the legend to the bottom (1)
```

This is a good place to start when doing exploratory analysis, but it's better
to use more sophisticated functions to produce better-looking plots.


## ggplot
ggplot is the tool of choice for general X-Y plots. It is possible to 
```{r, cache=TRUE}
library(ggplot2)  # Load the ggplot library
library(viridis)  # Load the library for a nice colour palette
ggplot(mapa.okres.sf) +
  geom_sf(aes(fill = plocha_uzemia)) +
  scale_fill_viridis("Plocha m\u00B2") +
  ggtitle("Plocha okresov") +
  theme_bw()
```

Whatever applied to ggplot craft, applies here as well.


## tmap

This is a very versatile library that allows integration with Open Street Map
data and interactive maps on top of regular plotting. 

Before we proceed, we install the libraries and load them.

```{r, eval=FALSE, echo=TRUE}
install.packages('tmap')
```

```{r, echo=TRUE, cache=TRUE, results='hide'}
library(tmap)
```

```{r, cache=TRUE, echo=TRUE}
tmap_mode("plot")
tm_shape(mapa.okres.sf) +
  tm_fill(col = "plocha_uzemia", style = "cont", fill.title="Plocha m2", scale = 0.8, alpha = 0.7) + 
  tm_style('gray', title = 'Plocha okresov m\u00B2')
```

For example, we can do a breakdown by kraj.

```{r, cache=TRUE, echo=TRUE}
tmap_mode("plot")
tm_shape(mapa.okres.sf) +
  tm_fill(col = "plocha_uzemia", style = "cont", fill.title="Plocha m2", scale = 0.8, alpha = 0.7) + 
  tm_style('gray', title = 'Plocha okresov m\u00B2') + 
  tm_facets('kraj')
```


For plots with fewer components (e.g. kraj), we might want to add labels, for
example:

```{r, cache=TRUE, echo=TRUE}
tmap_mode("plot")
tm_shape(mapa.kraj.sf) +
  tm_fill(col = "plocha_uzemia", style = "cont", fill.title="Plocha m2", scale = 0.8, alpha = 0.7) + 
  tm_style('gray', title = 'Plocha krajov m\u00B2') + 
  tm_text('kraj')
```

To save the map, use `tmap_save()` function.

```{r, eval=F}
mapa.okres.sf.gps <- st_transform(x = mapa.okres.sf, crs=4326)
osm_tiles = tmaptools::read_osm(st_bbox(mapa.okres.sf.gps), type = "osm")
tmap.object <- qtm(osm_tiles, raster.alpha = 0.6) + 
  tm_shape(mapa.okres.sf.gps) +
  tm_fill(col = "Shape_Area", style = "cont", fill.title="Plocha m2", scale = 0.8, alpha = 0.7) + 
  tm_style('gray', title = 'Plocha okresov')
```


## Interactive maps

```{r interactive_map, cache=TRUE, echo=TRUE}
tmap_mode("view")
tm_shape(mapa.okres.sf) +
  tm_fill(col = "plocha_uzemia", style = "cont", fill.title="Plocha m2", scale = 0.8, alpha = 0.7) + 
  tm_style('gray', title = 'Plocha okresov m\u00B2')
```

The interactive maps can be saved in the `html` format and can be easily
shared. The background maps are loaded dynamically from the internet.


# Operations


## Subsetting


## Distance metrics

## Working with different geometries
Working with grids and the administrative districts.


# Case study 1: Prezidentske volby, prve kolo.

1. Download the raw dataset of the results from [here](http://volby.statistics.sk/prez/prez2019/sk/download.html).
2. Load the results, merge it with the geography you are interested in (obec,
   kraj)
3. Plot the results.


# Case study 2: Unemployment rate.




